---
title: "Data analysis project 12"
format: html
editor: visual
execute: 
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(dplyr)
library(ggplot2) 
library(tidyverse)
library(tidymodels)
library(performance)
library(tinytex)

obesity <- read.csv("DAProject12.csv")
BMI <- obesity
```

# Introduction {#sec-intro}

Over the years, there has been an increase in the health awareness of the Scottish population, with a fear that the country as a whole is becoming overweight. The body mass index (BMI) is a quick way to measure weight relative to height, and screen for people that are underweight, overweight, or obese.

We aim to answer two questions:

1.  Has the BMI in Scotland changed over the given years of the Scottish Health Survey?

2.  Are there any difference in the BMI distributions by age, gender, socio-economic status or lifestyle factors?

# Exploratory data analysis {#sec-expdata}

>>>>>>> Formal-Analysis-Q2
```{r}

bmi_trends <- obesity%>% group_by(Year, BMIgroup) %>% summarise(count = n(), .groups =
"drop") %>% group_by(Year) %>% mutate(proportion = count / sum(count))

```
>>>>>>> Formal-Analysis-Q2

# Formal data analysis {#sec-formdata}

| Year | Obese proportions |
|------|-------------------|
| 2013 | 0.30300569        |
| 2014 | 0.30181715        |
| 2015 | 0.30645161        |
| 2016 | 0.30536494        |

```{r}
bmi_trends <- bmi_trends %>%
  mutate(obese_binary = ifelse(BMIgroup == "Obese", 1, 0))

```

```{r}
#| echo: false
bmi_trends$BMIgroup<-as.factor(bmi_trends$BMIgroup)
levels(bmi_trends$BMIgroup)<-c("Not Obese","Obese","Not Obese","Not Obese")
trend_table <- table(bmi_trends$Year, bmi_trends$obese_binary)
head(trend_table)
prop.trend.test(x = trend_table[,2],  
                n = rowSums(trend_table))  
unique(bmi_trends$Year)
unique(bmi_trends$obese_binary)
bmi_trends$Year<-as.factor(bmi_trends$Year)
chisq.test(trend_table)
```

```{r}
glm_model <- glm(obese_binary ~ as.numeric(Year), 
                 data = bmi_trends, family = binomial)
summary(glm_model)

```
### Question 2

To answer the second question, we use linear regression to find out if age, gender, socio-economic status, or lifestyle factors are significant predictors for obesity.

We start by looking at if age is a significant predictor.

```{r}
#| echo: false
#| eval: true
BMI$BMIgroup <- as.factor(BMI$BMIgroup)
levels(BMI$BMIgroup) <- c("Not Obese", "Obese", "Not Obese", "Not Obese")

# Regression to check if Age is a significant predictor
model_age <- glm(BMIgroup ~ Age,
                     data = BMI,
                     family = binomial)

# Interpretation is in log-odds!
model_age %>% tidy(conf.int = T)

# Model diagnostics - Checking for overdispersion
check_overdispersion(model_age)
```

We now look at sex to consider if it is significant

```{r}
#| echo: false
#| eval: true
# Regression to check if Sex is a significant predictor
model_sex <- glm(BMIgroup ~ Sex,
                     data = BMI,
                     family = binomial)

# Interpretation is in the log-odds scale
model_sex %>% tidy(conf.int = T) # P-value of 0.0247 -> sex is significant

# Then checking for overdispersion
check_overdispersion(model_sex)
```

We now look at education to consider if there is a difference in obesity based on socio-economic factors

```{r}
#| echo: false
#| eval: true
# Regression to check if the various levels of education are significant as predictors
model_edu <- glm(BMIgroup ~ Education,
                     data = BMI,
                     family = binomial)

# interpretation is in the log-odds scale
model_edu %>% tidy(conf.int = T) # All are significant

# We again check for overdispersion
check_overdispersion(model_edu)
```

## Results {#sec-results}

# Conclusions {#sec-conc}
>>>>>>> Formal-Analysis-Q2
