---
title: "Data analysis project 12"
format: html
editor: visual
execute: 
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(dplyr)
library(ggplot2) 
library(tidyverse)
library(tidymodels)
library(performance)
library(tinytex)
library(sjPlot)

obesity <- read.csv("DAProject12.csv")
BMI <- obesity
```

# Introduction {#sec-intro}

Over the years, there has been an increase in the health awareness of the Scottish population, with a fear that the country as a whole is becoming overweight. The body mass index (BMI) is a quick way to measure weight relative to height, and screen for people that are underweight, overweight, or obese.

We aim to answer two questions:

1.  Has the BMI in Scotland changed over the given years of the Scottish Health Survey?

2.  Are there any difference in the BMI distributions by age, gender, socio-economic status or lifestyle factors?

# Exploratory data analysis {#sec-expdata}

>>>>>>> Formal-Analysis-Q2
```{r}

bmi_trends <- obesity%>% group_by(Year, BMIgroup) %>% summarise(count = n(), .groups =
"drop") %>% group_by(Year) %>% mutate(proportion = count / sum(count))

```
>>>>>>> Formal-Analysis-Q2

# Formal data analysis {#sec-formdata}

| Year | Obese proportions |
|------|-------------------|
| 2013 | 0.30300569        |
| 2014 | 0.30181715        |
| 2015 | 0.30645161        |
| 2016 | 0.30536494        |

```{r}
bmi_trends <- bmi_trends %>%
  mutate(obese_binary = ifelse(BMIgroup == "Obese", 1, 0))

```

```{r}
#| echo: false
bmi_trends$BMIgroup<-as.factor(bmi_trends$BMIgroup)
levels(bmi_trends$BMIgroup)<-c("Not Obese","Obese","Not Obese","Not Obese")
trend_table <- table(bmi_trends$Year, bmi_trends$obese_binary)
head(trend_table)
prop.trend.test(x = trend_table[,2],  
                n = rowSums(trend_table))  
unique(bmi_trends$Year)
unique(bmi_trends$obese_binary)
bmi_trends$Year<-as.factor(bmi_trends$Year)
chisq.test(trend_table)
```

```{r}
glm_model <- glm(obese_binary ~ as.numeric(Year), 
                 data = bmi_trends, family = binomial)
summary(glm_model)

```
### Question 2

To answer the second question, we use linear regression to find out if age, gender, socio-economic status, or lifestyle factors are significant predictors for obesity.

We start by looking at if age is a significant predictor for a persons BMI group. We are fitting the following linear model: 
$$
y_i = \beta_0 + \beta_1 x_i + \epsilon_i, ~~~~~ \epsilon_i \sim N(0, \sigma^2), ~~~~~ i = 1, ... , 14017
$$

Here, $y_i$ is the BMI group of the ith individual, $x_i$ is the age of the ith individual, and $\beta_0$, $\beta_1$ are regression coefficients

```{r}
#| echo: false
#| eval: true
#| tab-cap: output from linear regression with age as the predictor
#| label: tab-regression1
#| warnings: false
#| message: false


BMI$BMIgroup <- as.factor(BMI$BMIgroup)
levels(BMI$BMIgroup) <- c("Not Obese", "Obese", "Not Obese", "Not Obese")
BMI$BMIgroup <- relevel(BMI$BMIgroup, ref = "Obese")


# Regression to check if Age is a significant predictor
model_age <- glm(BMIgroup ~ Age,
                     data = BMI,
                     family = binomial)

tab_model(model_age, show.r2 = FALSE, show.obs = FALSE)
```
As seen from the p-value, we can infer that age is a significant predictor for a persons BMI group. As someones age increases by one year, the log odds of the person being obese increase by a factor of 0.99. 

We now check the assumptions of our model, to find out if the model is valid.

``` {r}
#| label: fig-diagnostics1
#| fig-cap: Diagnostic plots for the age model

# Model diagnostics
check_model(model_age, panel = TRUE)
```
There is no issues that can be seen in the posterior predictive check, influential obesvations graph, and the uniformity of resuduals graph. There does appear to be a large amount of points falling outiwth the error bounds in the binned residuals graph. We would expect about 95% of points to fall within the error bounds if the model is true. As there is a large amount of points plotted it is hard to tell if there is 95% within the error bounds, as there may be overlap. However, it appears to be good enough. 

We now look at sex to consider if it is significant through fitting the following model:

$$
y_i = \beta_0 + \beta_1 x_i + \epsilon_i, ~~~~~ \epsilon_i \sim N(0, \sigma^2), ~~~~~ i = 1, ... , 14017
$$

Where $x_i$ is an indicator variable taking the value 1 if the individual is a male and the value 0 if the individual is a female.

```{r}
#| echo: false
#| eval: true
#| tab-cap: output from linear regression with sex as the predictor
#| label: tab-regression2
#| warnings: false
#| message: false

# Regression to check if Sex is a significant predictor
model_sex <- glm(BMIgroup ~ Sex,
                     data = BMI,
                     family = binomial)

tab_model(model_sex, show.obs = FALSE)
```

As seen from table 2, the p-value indicates that sex is a significant predictor at the 5% significance level. Furthermore, the log odds of being obese increase by  1.12 if the individual is a male.

Again, we check our model assumptions to find out if our results are valid.

```{r}
#| label: fig-diagnostics2
#| fig-cap: Diagnostic plots for the sex model

# Model diagnostics
check_model(model_sex, panel = TRUE)
```
The plots in figure 2 look good enough. There may be issues with the binned residuals plot, however the issue likely stems from sex being a factor with 2 levels giving us only 2 points.

We now look at education to consider if there is a difference in obesity based on socio-economic factors. We fit the following model

```{r}
#| echo: false
#| eval: true
#| tab-cap: output from linear regression with education as the predictor
#| label: tab-regression3
#| warnings: false
#| message: false

# Regression to check if the various levels of education are significant as predictors
model_edu <- glm(BMIgroup ~ Education,
                     data = BMI,
                     family = binomial)

tab_model(model_edu, show.obs = F)
```

As seen from table 3, all the levels of education are considered significant predictors of the persons BMI group. The log odds of a person being obese is 3.05 for a person with a degree, then increases by 0.53-0.84 depending on qualification level. 

Again, we have to check the model assumptions. 

```{r}
#| label: fig-diagnostics3
#| fig-cap: Diagnostic plots for the education model

# Model diagnostics
check_model(model_edu, panel = TRUE)
```

The interpretation of figure 3 is much like the previous figures. The only graph that leaves cause for concern is the binned residuals graph, but again, becuase it is a factor, it is always going to not work as well as a continuous variable. Thus, the model is appropriate.

## Results {#sec-results}

# Conclusions {#sec-conc}


>>>>>>> Formal-Analysis-Q2

From the previous tests, we have seen that age, sex, and education are all significant in the prediction of an individuals BMI group, indicating differences in obesity across these factors. As age increases, the log odds of obesity rise. Males have higher log odds of being obese than females, and lower education levels are associated with greater log odds of obesity.