---
title: "TBD"
format: pdf
editor: visual
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(tidymodels)
library(performance)
library(tinytex)

BMI <- read.csv("DAProject12.csv")
```

## Introduction

Over the years, there has been an increase in the health awareness of the Scottish population, with a fear that the country as a whole is becoming overweight. The body mass index (BMI) is a quick way to measure weight relative to height, and screen for people that are underweight, overweight, or obese.

We aim to answer two questions:

1.  Has the BMI in Scotland changed over the given years of the Scottish Health Survey?

2.  Are there any difference in the BMI distributions by age, gender, socio-economic status or lifestyle factors?

## Exploratory Analysis

## Formal Analysis

### Question 2

To answer the second question, we use linear regression to find out if age, gender, socio-economic status, or lifestyle factors are significant predictors for obesity.

We start by looking at if age is a significant predictor.

```{r}
#| echo: false
#| eval: true
BMI$BMIgroup <- as.factor(BMI$BMIgroup)
levels(BMI$BMIgroup) <- c("Not Obese", "Obese", "Not Obese", "Not Obese")

# Regression to check if Age is a significant predictor
model_age <- glm(BMIgroup ~ Age,
                     data = BMI,
                     family = binomial)

# Interpretation is in log-odds!
model_age %>% tidy(conf.int = T)

# Model diagnostics - Checking for overdispersion
check_overdispersion(model_age)
```

We now look at sex to consider if it is significant

```{r}
#| echo: false
#| eval: true
# Regression to check if Sex is a significant predictor
model_sex <- glm(BMIgroup ~ Sex,
                     data = BMI,
                     family = binomial)

# Interpretation is in the log-odds scale
model_sex %>% tidy(conf.int = T) # P-value of 0.0247 -> sex is significant

# Then checking for overdispersion
check_overdispersion(model_sex)
```

We now look at education to consider if there is a difference in obesity based on socio-economic factors

```{r}
#| echo: false
#| eval: true
# Regression to check if the various levels of education are significant as predictors
model_edu <- glm(BMIgroup ~ Education,
                     data = BMI,
                     family = binomial)

# interpretation is in the log-odds scale
model_edu %>% tidy(conf.int = T) # All are significant

# We again check for overdispersion
check_overdispersion(model_edu)
```

## Conclusion
