---
title: "Data analysis project 12"
format: html
editor: visual
execute: 
  echo: false
---

```{r}
#| warning: false
#| message: false

library(ggplot2)
library(tidyverse)
library(tidymodels)
library(gt)
library(MASS)
library(patchwork)
library(moderndive)
library(dplyr)
library(gapminder)
library(sjPlot)
library(stats)
library(readr)
library(janitor)
library(ggstats)
library(skimr)
library(knitr)
```

```{r}
#| warning: false
#| message: false

data1 <- read.csv("D:/1332/DAProject12.csv")

```
# Introduction {#sec-intro}
Obesity is a growing public health concern worldwide, and Scotland is no exception. With lifestyle, socio-economic, and demographic factors playing crucial roles in obesity prevalence, it is important to analyze trends and patterns over time. This study utilizes data from the Scottish Health Surveys (2013-2016) to explore the distribution of obesity and its relationship with age, sex, education, and other lifestyle factors.

The primary objectives of this analysis are:

1.To examine how the prevalence of obesity has changed from 2013 to 2016.

2.To investigate whether there are differences in obesity rates based on age, gender, socio-economic status, or educational attainment.

3.By understanding these trends, policymakers and health professionals can better address obesity-related health issues and develop targeted interventions to promote healthier lifestyles in Scotland.

# Exploratory data analysis {#sec-expdata}

These statistics can be illustrated in @fig-bar1 below. We can also see that there are no significant change in the proportion of obese individuals between 2013 and 2016.

```{r}
#| echo: false
#| warning: false
#| fig-cap: bar plot of BMIgroup by year
#| label: fig-bar1
#| fig-height: 2
ggplot(data1, aes(x = factor(Year), fill = BMIgroup)) +
  geom_bar(position = "fill") +
  labs(title = "BMI Group Proportion by Year", x = "Year", y = "Proportion") +
  theme_minimal()
```

Next, we want to explore the relationship between Age and BMIgroup.In @fig-box1 shows that the age distribution varies significantly across different BMI groups. Generally, obese individuals tend to be older, whereas the underweight group is comparatively younger. This indicates age is an important factor associated with BMI categories.

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Box plot showing the relationship between Age and BMIgroup"
#| label: fig-box1

boxplot(Age ~ BMIgroup, data = data1,
        main = "Age Distribution by BMI Group",
        xlab = "BMI Group",
        ylab = "Age")

```
Next, we analyse the relationship between BMI Group Proportion by Sex. In @fig-bar2 we see there are noticeable differences in BMI group proportions between males and females. Males appear slightly more likely to be overweight, whereas females show relatively higher proportions in the normal and obese categories. Gender thus influences BMI distribution patterns.
```{r}
#| echo: false
#| warning: false
#| fig-cap: Bar plot showing the relationship between Sex and BMIgroup
#| label: fig-bar2
#| fig-height: 2
ggplot(data1, aes(x = Sex, fill = BMIgroup)) +
  geom_bar(position = "fill") +
  labs(title = "BMI Group Proportion by Sex", x = "Sex", y = "Proportion") +
  theme_minimal()
```
Finally, we want to assess the relationship between Education and BMIgroup. In @fig-bar3 shows that the education level is related to BMI groups. Individuals with higher educational qualifications generally have higher proportions of normal weight and lower proportions of obesity. Conversely, individuals with fewer or no qualifications show a greater proportion of obesity. Socio-economic factors, represented by education levels, clearly impact obesity rates.
```{r}
#| echo: false
#| warning: false
#| fig-cap: Bar plot showing the relationship between Education and BMIgroup
#| label: fig-bar3
#| fig-height: 2
ggplot(data1, aes(x = Education, fill = BMIgroup)) +
  geom_bar(position = "fill") +
  labs(title = "BMI Group Proportion by Education", x = "Education", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```